Neural correlates of the deployment of spatial attention, and their modulation by repetitive movements

### 2023-09-14 manuscript.qmd
Added references file

---
title: "Stimming paper"
author: "Cameron Smith & Daniel H. Baker"
execute: 
  echo: false
  include: false
  output: false
format: pdf
---

@@

The ability to orient and control the deployment of our attention to spatial locations or features of our environment is invaluable to our ability to perform tasks, focusing on relevant stimuli and disregarding those that are unnecessary for the task at hand (Knudson, 2007; Hopfinger et al., 2000; Vohs and Baumeister, 2013). A classic technique for studying the deployment of attention in space is the Posner cueing paradigm, which demonstrates that it is possible for attention to be deployed to spatial locations independently of eye movements (Posner, 1980).

Individuals with neurodevelopmental disorders such as attention deficit hyperactivity disorder (ADHD) and autism present with deficits in attention (American Psychiatric Association, 2013; Allen and Courchesne, 2001). It has been suggested that repetitive stereotypies, or ‘stimming’ behaviours, are linked to these deficits in attention- in the few studies where they have been asked, individuals who stim report that it benefits their ability to focus, and that the behaviours aid in emotional regulation (Kapp et al., 2019; Steward, 2015). It has been proposed that an improved ability to focus through stimming may have a neural basis (McCarty and Brumback, 2021). Building on these theories and research, we tested the effect of stimming on performance in a Posner cueing task.

The Posner cueing paradigm (Posner, 1980) involves the participant attending to a central fixation point, then receiving a cue (which may be exogenous, where attention is captured through bottom-up processes; or endogenous, where attention is controlled by top-down processes, such as attending selectively to cues of the correct colour (Leblanc et al., 2008)) indicating that a stimulus is about to appear either to the left or the right of the fixation point (Thiery et al., 2016). The assumption underlying the task is that if participants successfully orient their attention to the cued location, then their reaction times will be shorter for valid trials (where the stimulus appears at the cued location) than for invalid trials (where the stimulus appears at the uncued location); this has been demonstrated extensively (Thiery et al., 2016). This reaction time difference was found by Posner et al. (1978, as cited in Johnson and Haggard, 2003) to be 25ms faster for valid trials and 40ms slower for invalid trials (compared to a neutral cue condition).

@@

A number of developmental disorders involve deficits in attention, notably ADHD and autism. In the case of ADHD, attention deficits are a defining feature of the condition (American Psychological Association, 2013). For autism, attention deficits are not an explicit part of the diagnostic criteria (although restricted interests, which are included, are somewhat related), but both attentional deficits and strengths are widely reported in autistic individuals. They are typically more distractible, demonstrating ‘underselective’ attention, but also demonstrate ‘overselective’ attention where they attend intensely to a more limited subset of stimuli. This ‘overselective’ attention has been suggested to be linked to restricted interests, which is also sometimes referred to as ‘hyperfocus’, where a task is the subject of intense focus for the individual (e.g., Dupuis et al., 2022). These differences appear to have a neural basis- they have been linked to abnormally wide sulci in the parietal lobe (Allen and Courchesne, 2001). Hyperfocus is also widely self-reported in ADHD, despite a lack of clinical research into the phenomenon (Hupfeld et al., 2019).

@@

Current research into stimming places a large focus on designing interventions to eliminate the behaviours, for example the use of weighted vests (Fertel-Daly et al., 2001). Taking this approach without first attempting to understand if there are benefits to the behaviour seems ill-advised. McCarty and Brumback (2021), for example, proposed based on self-reports of the benefits of stimming (focus, coping with overwhelming sensory input, and relaxation) that regular motor movements may generate (or be a by-product of) rhythmic oscillations in the motor cortex, which entrain oscillations in the sensory cortex. This entrainment might improve information transfer in the sensory cortex, effectively normalising the atypical motor and sensory oscillations observed in the brains of autistic individuals both during activity (Milne et al., 2009; Snijders et al., 2013; Murphy and Benítez-Burraco, 2017; as cited in McCarty and Brumback) and at rest (Wang et al., 2013; Berman et al., 2015; as cited in McCarty and Brumback).

>>>>

---
title: "Neural correlates of the deployment of spatial attention, and their modulation by repetitive movements"
author: "Cameron Smith & Daniel H. Baker"
bibliography: references.bib
csl: elife.csl
execute: 
  echo: false
  include: false
  output: false
format: pdf
---

@@

The ability to orient and control the deployment of our attention to spatial locations or features of our environment is invaluable to our ability to perform tasks, focusing on relevant stimuli and disregarding those that are unnecessary for the task at hand (Knudson, 2007; Hopfinger et al., 2000; Vohs and Baumeister, 2013). A classic technique for studying the deployment of attention in space is the Posner cueing paradigm, which demonstrates that it is possible for attention to be deployed to spatial locations independently of eye movements [@Posner1980].

Individuals with neurodevelopmental disorders such as attention deficit hyperactivity disorder (ADHD) and autism present with deficits in attention (American Psychiatric Association, 2013; Allen and Courchesne, 2001). It has been suggested that repetitive stereotypies, or ‘stimming’ behaviours, are linked to these deficits in attention -- in the few studies where they have been asked, individuals who stim report that it benefits their ability to focus, and that the behaviours aid in emotional regulation (Kapp et al., 2019; Steward, 2015). It has been proposed that an improved ability to focus through stimming may have a neural basis (McCarty and Brumback, 2021). Building on these theories and research, we tested the effect of stimming on performance in a Posner cueing task.

The Posner cueing paradigm [@Posner1980] involves the participant attending to a central fixation point, then receiving a cue (which may be exogenous, where attention is captured through bottom-up processes; or endogenous, where attention is controlled by top-down processes, such as attending selectively to cues of the correct colour (Leblanc et al., 2008)) indicating that a stimulus is about to appear either to the left or the right of the fixation point (Thiery et al., 2016). The assumption underlying the task is that if participants successfully orient their attention to the cued location, then their reaction times will be shorter for valid trials (where the stimulus appears at the cued location) than for invalid trials (where the stimulus appears at the uncued location); this has been demonstrated extensively (Thiery et al., 2016). This reaction time difference was found by Posner et al. (1978, as cited in Johnson and Haggard, 2003) to be 25ms faster for valid trials and 40ms slower for invalid trials (compared to a neutral cue condition).

@@

A number of developmental disorders involve deficits in attention, notably ADHD and autism. In the case of ADHD, attention deficits are a defining feature of the condition (American Psychological Association, 2013). For autism, attention deficits are not an explicit part of the diagnostic criteria (although restricted interests, which are included, are somewhat related), but both attentional deficits and strengths are widely reported in autistic individuals. They are typically more distractible, demonstrating ‘underselective’ attention, but also demonstrate ‘overselective’ attention where they attend intensely to a more limited subset of stimuli. This ‘overselective’ attention has been suggested to be linked to restricted interests, which is also sometimes referred to as ‘hyperfocus’, where a task is the subject of intense focus for the individual (e.g., Dupuis et al., 2022). These differences appear to have a neural basis -- they have been linked to abnormally wide sulci in the parietal lobe (Allen and Courchesne, 2001). Hyperfocus is also widely self-reported in ADHD, despite a lack of clinical research into the phenomenon (Hupfeld et al., 2019).

@@

Current research into stimming is largely focussed on designing interventions to eliminate the behaviours, for example the use of weighted vests (Fertel-Daly et al., 2001). Taking this approach without first attempting to understand if there are benefits to the behaviour seems ill-advised. McCarty and Brumback (2021), for example, proposed based on self-reports of the benefits of stimming (focus, coping with overwhelming sensory input, and relaxation) that regular motor movements may generate (or be a by-product of) rhythmic oscillations in the motor cortex, which entrain oscillations in the sensory cortex. This entrainment might improve information transfer in the sensory cortex, effectively normalising the atypical motor and sensory oscillations observed in the brains of autistic individuals both during activity (Milne et al., 2009; Snijders et al., 2013; Murphy and Benítez-Burraco, 2017; as cited in McCarty and Brumback) and at rest (Wang et al., 2013; Berman et al., 2015; as cited in McCarty and Brumback).
